Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-07-05 patchy-selection-paper.tex
peter edited through entirely

%%%%%%%%%%%
\abstract{
Populations within a species often repeatedly locally adapt to similar environmental conditions
across their range. Such adaptation can occur through convergent
evolution as different alleles arise and spread in different parts of
the range. Or it can occur through the spread of alleles by migration
acting to synchronizing adaptation across the species.
The tension between the two reflects the degree of 
constraint imposed on evolution by the underlying genetic
architecture versus how how effectively selection acts to inhibt the geographic spread of local
adapted alleles.
This paper studies a model of the the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a new, locally adapted allele to appear in a
patch of favorable habitat through new mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of patch separation?
How can we tell which has occurred, i.e. what population genetic
signal is left by the spread of migrant alleles and how long does this
signal persist for?
To answer these questions we decompose the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
This provides a set of tractable approximations to understand the role
of geographic separation between favorable patches in promoting
convergent adaptation and the genomic signals left.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

@@

% Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
% sometimes over deep time scales \citep{deephomologypapers}. 
% Convergence on a molecular level also suggests that the path 
% accessible to adaptation is sometimes relatively constrained \citep{stern2013genetic}, 
% as would be the case if only a few changes confer the advantagous phenotype 
% and are sufficiently free of deleterious pleiotropic consequences.

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testiment to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic,zhen2012parallel,martin2013repeated}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetics changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \citep{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

%\citep{wright2013indirect}
One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{macnair1991evolution,schat1996identical,turner2010serpentine};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches.
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes
\citep{benson1933concealing, dice1940ecologic}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is on a number of black lava flows has much darker pelage than on intervening,
much lighter soil \citep{DiceBlossom1937}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible for the dark pigmentation found on different outcrops \citep{hoekstra2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}
and in lizards \citep{rosenblum2010molecular}.

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?
How easy will it be to detect adaptive alleles that are shared by
migration, i.e. over what genomic scale will a population genetic
signal be visible?

We work in a model of continuous geography,
using a variety of known results and new methods.
In section \ref{ss:patchymutation} we develop a simple approximation,
eqn. \eqref{eqn:mutrate}, for the rate at which patches become
established by new mutations. Then in section \ref{ss:patchymigration} we
develop an approximation, eqn. \eqref{eqn:migrate}, to the rate which alleles move from an
existing patch to colonize a novel patch (this represents a big chunk
of the results of the paper). We combine these two results in section
\ref{ss:probparallel} to develop an approximation to the probability of
parallel adaptation, eqn. \eqref{eqn:parallel_prob}.
To understand the genomic signal left by adaptations shared by
migration in section \ref{ss:haplotype_length} we approximate the time
it will take for an allele to transit between patches,
eqn. \eqref{eqn:mean_tau}, and use this to approximate the length of
haplotype that hitchhikes with it,
eqn. \eqref{eqn:haplotype_length}. We then quantify how quickly this
haplotype decays within patches eqn. \eqref{eqn:rec_rate_within} and
use this to obtain properties about the equilibrium distribution of
haplotype lengths
\eqref{eqn:within_haplotype_length}  in section \ref{ss:patch_haplotype}. 
Finally in section \ref{ss:applications} we apply this work to understand the geographic scale of
convergent evolution in \textit{Chaetodipus intermedius}.

% http://www.nature.com/hdy/journal/v94/n2/full/6800600a.html   -- this is \citep{hoekstra2004local}

%and that a single mutational change was sufficient to adapt populations to the novel environment. 
%We assumed that there was no standing variation for the adaptive allele, such that parallel mutation must be due to multiple mutations after the environmental shift. In that setting we found that there was a characteristic geographic scale, over which we would expect multiple instances of our adaptive allele to have arisen in parallel. This characteristic length could be expressed in terms of a simple compound parameter determined by our parameters of interest.
%Recall questions of origin of adaptations

%References on parallel adapation, standing variation, etc.:
 % coop,
 % pennings and hermisson.
%Note pennings and hermisson found standing deleterious variation was very important.

%recall peromyscus examples with patchy environments

%Summarize previous paper

%Question assumptions of no standing variation and homogenous selection

@@

We are interested in the establishment of mutations in the ``poor'' patches by either
migration or mutation, and so are mainly interested in whether the allele
can escape initial loss by drift when rare. 
Therefore, we do not have to specify the the fitness of the homozygote; 
only that the dynamics of the allele when rare 
are determined by the fitness of the heterozygote. 
More general dominance will only make small corrections to the the dynamics until initial fixation,
with the exception of the recessive case, which we omit.
For the most part, 
we follow the literature in treating the diploid model as essentially haploid.

For the sake of simplicity, we assume a constant haploid population density $\rho$, 
across both types of habitat,
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). Differences in population
densities on and off the patches may arise if selection is hard, in
that case these differences can be incorporated into the density \citep[see][]{lenormand2002limits}. 
%We also aim to describe the scenario where it is sensible to think of large, discrete patches;
%the case of a fine-scale heterogeneous environment is treated
%\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATIOn}

\paragraph{Past work on spatial models}
We will make use of several previous results from the literature on
migration in spatially varying selection models. 
\citet{haldane1948theory}, \citet{fisher1950frequencies}, and later \citet{slatkin1973geneflow} first analyzed
deterministic models of of this sort.
\citet{nagylaki1975conditions} and \citet{conley1975application} showed that in one dimension, if the physical width of the patch is less than $(\sigma/\sqrt{2s_p}) \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,  % see p.603 in Nagylaki
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed  
that patches must be at least this critical size so that a new mutation has positive probability of establishment
(in the large population density limit).
The same paper also showed that
this probability of establishment of a new mutation decays exponentially with distance from the patch, 
with the scale given by $\sigma/\sqrt{s_m}$. Mutations appearing within the patch have probability of establishment
strictly less than the probability for a panimictic population,
but approaching this as the size of the patch increases.
This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane1927mathematical,fisher1930genetical}.
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}. 
More general work on the mathematically equivalent problem of density-dependent population growth 
has obtained much more general criteria
under which a habitat configuration will support a stable polymorphic equilibrium \citep{cantrell1989diffusive},
and determined the habitat configuration that maximizes the probability of establishment \citep{lou2006minimization}.

@@

Consider first a single, isolated poor habitat patch of area $A$ in which no $B$ allele has yet become established. 
We first compute the time scale on which new $B$ mutations appear and fix in such a patch.
As we are interested in patches where local adaptation can occur,
we will assume that our patch is larger than the cutoff for local establishment 
mentioned above.

Let $p(x)$ be the probability that a new mutant $B$ allele arising at location $x$
relative to the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approxmation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to assume that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
We would then approximate the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is symmetric about the edge of the patch.
We examine this more generally via exact calculation of $p(x)$ in Appendix \ref{apx:establishment_sims}.

@@

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, we dissect the process by which an allele transits between the patches,
in the process obtaining other useful information about the genealogy of $B$ alleles.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at location $x$ relative to the patch by $q(x)$,
and assume that the shape of the patch is at least roughly circular.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
and show that $q(x)$ decays exponentially for large $|x|$,
with a square-root correction in two dimensions:
\begin{align} \label{eqn:eqfreq}
  q(x) \approx C \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), 
and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients.
These asymptotics fit simulations quite well, as shown in Figure~\ref{fig:sim_occupation_freqs}.
To be clear about the assumptions (especially regarding scaling) implicit here,
we provide a derivation in Appendix~\ref{apx:eqfreq}; 
as well as justification for the asymptotics in Appendix \ref{apx:asymptotics}.
In one dimension, the equation can be solved to give $q(x)$ 
in terms of Jacobi elliptic functions; see Appendix \ref{apx:elliptic_integrals}.
In applications we fit $C$ to the data;
to get concrete numbers from theory we take $C=1$ if necessary.
From simulations, $C$ seems to be not too far from 1 in practice (see below).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \caption{
  \textbf{(A)} 
  Mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  a one-dimensional grid of 201 demes, and
  a two-dimensional grid of $101\times 101$ demes.
  Superimposed in black is the decay predicted by \eqref{eqn:eqfreq},
  with $C$ chosen arbitrarily to give a reasonable fit.
  (The disagreement at long distances is due to boundary effects near the edge of the range.)
  \textbf{(B)} 
  Fluctuations: colored lines are snapshots of allele frequencies at evenly spaced times
  from a one-dimensional grid of 151 demes;
  the solid lines shows the cumulative mean occupation frequency across all generations.
  For both plots, the simulation was run for 1000 generations;
  the focal allele is beneficial in the central 20 demes, with $s_p=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$; 
  demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
   }   \label{fig:sim_occupation_freqs}
\end{figure}

This expected frequency $q(x)$ is the \emph{time-averaged} occupation frequency,
or in other words, the total number of $B$ alleles found across $T$ generations near location $x$, per unit area, divided by $T$.
If, for instance, $q(x)=.01$ and the population density is $\rho=10$ individuals per unit area, 
then in a survey tract of unit area around $x$ we expect to find one individual every 10 generations
-- or, perhaps more likely, 10 individuals every 100 generations.
This points to an important fact: 
close to the original patch, the ``equilibrium frequency'' $q(x)$ describes well the density of the $B$ allele at most times,
but far away from the patch, 
the equilibrium is composed of rare excursions of families of $B$
alleles, interspersed by long periods of absence. 
An example of one of these rare excursions is shown in Figure~\ref{fig:sim_snapshots}.
The relevant time scale on which $B$ alleles migrate between patches is given by the rate of appearance of such families.
To compute the rate of establishment by such migration,
we argue that these families (suitably defined) can be approximated by
independent branching processes whose spatial motion is described by Brownian motion.
Properties of these, combined with equation \eqref{eqn:eqfreq},
will lead us eventually to the expression \eqref{eqn:migrate} for the rate of effective migration,
an approximation that is valid if the distance $R$ is large (at least five times $\sigma/\sqrt{2s_m}$, say).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-snapshots}
  \end{center}
  \caption{
  The temporal dynamics of the same simulation as Figure~\ref{fig:sim_occupation_freqs}B.
  A segment of the temporal dynamics: space is again shown across the horizontal axis;
  with time on the vertical axis; demes are colored darker red the greater then allele frequency at the corresponding time.
  The 50\% and 5\% frequency contour lines are in grey,
  and the genealogy of 50 individuals in a rare long-distance excursion is traced back with black lines.
  }   \label{fig:sim_snapshots}
\end{figure}

>>>>

%%%%%%%%%%%
\abstract{
Species often encounter, and adapt to, many patches of locally similar environmental conditions
across their range. 
Such adaptation can occur through convergent evolution as different alleles arise and spread in different parts of the range, 
or through the spread of alleles by migration
acting to synchronize adaptation across the species.
The tension between the two reflects the degree of 
constraint imposed on evolution by the underlying genetic
architecture versus how how effectively selection acts to inhibit the geographic spread of locally
adapted alleles.
This paper studies a model of the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a new, locally adapted allele to appear in a
patch of habitat where it is favored through new mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of distance between the patches?
How can we tell which has occurred, i.e.\ what population genetic
signal is left by the spread of migrant alleles and how long does this
signal persist for?
To answer these questions we decompose the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
This provides a way to understand the role of geographic separation between patches 
in promoting convergent adaptation and the genomic signals it leaves behind.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

@@

% Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
% sometimes over deep time scales \citep{deephomologypapers}. 
% Convergence on a molecular level also suggests that the path 
% accessible to adaptation is sometimes relatively constrained \citep{stern2013genetic}, 
% as would be the case if only a few changes confer the advantageous phenotype 
% and are sufficiently free of deleterious pleiotropic consequences.

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testament to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic,zhen2012parallel,martin2013repeated}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \citep{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

%\citep{wright2013indirect}
One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{kruckeberg1951intraspecific,turner2010serpentine,macnair1991evolution,schat1996identical};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches.
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes
\citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is on a number of black lava flows has much darker pelage than on intervening,
much lighter soil \citep{DiceBlossom1937}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible for the dark pigmentation found on different outcrops \citep{hoekstra2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism,dice1940ecologic}
and in lizards \citep{rosenblum2010molecular}.

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?
How easy will it be to detect adaptive alleles that are shared by
migration, i.e.\ over what genomic scale will a population genetic
signal be visible?

We work in a model of continuous geography,
using a variety of known results and new methods.
In section \ref{ss:patchymutation} we develop a simple approximation,
equation \eqref{eqn:mutrate}, for the rate at which patches become established by new mutations. 
The most important conceptual work of the paper occurs in section \ref{ss:patchymigration},
where we develop an understanding of the process by which alleles move from an existing patch 
to colonize a novel patch, culminating in equation \eqref{eqn:migrate} for the rate at which this happens.
We combine these two results in section \ref{ss:probparallel} 
to discuss to the probability of parallel adaptation, equation \eqref{eqn:parallel_prob}.
To understand the genomic signal left by adaptations shared by
migration in section \ref{ss:haplotype_length} we approximate the time
it will take for an allele to transit between patches,
equation \eqref{eqn:mean_tau}, and use this to approximate the length of
haplotype that hitchhikes with it,
equation \eqref{eqn:haplotype_length}. We then quantify how quickly this
haplotype decays within patches equation \eqref{eqn:rec_rate_within} and
use this to obtain properties about the equilibrium distribution of
haplotype lengths
\eqref{eqn:within_haplotype_length}  in section \ref{ss:patch_haplotype}. 
Finally, in section \ref{ss:applications} we apply this work to understand the geographic scale of
convergent evolution in \textit{Chaetodipus intermedius}.

@@

We are interested in the establishment of mutations in the ``poor'' patches by either
migration or mutation, and so are mainly interested in whether the allele
can escape initial loss by drift when rare. 
Therefore, we do not have to specify the fitness of the homozygote; 
only that the dynamics of the allele when rare 
are determined by the fitness of the heterozygote. 
More general dominance will only make small corrections to the dynamics until initial fixation,
with the exception of the recessive case, which we omit.
In other words,
we follow the literature in treating the diploid model as essentially haploid.

Also, to avoid introducing more parameters,
we assume population density $\rho$ is equal across the range (even in the ``poor'' patches).
If this is not the case, in most equations below, the population density that should be used is the density
on the poor patches; see \citet{lenormand2002limits} for a more thorough approach.
We further assume that the variance in offspring number is $\xi^2$, 
and that the mean squared distance between parent and child is $\sigma^2$
(i.e.\ $\sigma$ is the dispersal distance). 
%We also aim to describe the scenario where it is sensible to think of large, discrete patches;
%the case of a fine-scale heterogeneous environment is treated
%\citep{elsewhere}. \gc{IS THIS LAST SENTENCE NEEDED IF SO GIVE CITATION} \plr{haven't found it yet}

\paragraph{Past work on spatial models}
We will make use of several previous results from the literature on
migration in spatially varying selection models. 
\citet{haldane1948theory}, \citet{fisher1950frequencies}, and later \citet{slatkin1973geneflow} first analyzed
deterministic models of of this sort.
\citet{nagylaki1975conditions} and \citet{conley1975application} showed that in one dimension, if the physical width of the patch is less than $(\sigma/\sqrt{2s_p}) \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,  % see p.603 in Nagylaki
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed  
that patches must be at least this critical size so that a new mutation has positive probability of establishment
(in the large population density limit).
The same paper also showed that
this probability of establishment of a new mutation decays exponentially with distance from the patch, 
with the scale given by $\sigma/\sqrt{s_m}$. Mutations appearing within the patch have probability of establishment
strictly less than the probability for a panmictic population,
but approaching this as the size of the patch increases.
This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane1927mathematical,fisher1930genetical}.
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}. 
More general work on the mathematically equivalent problem of density-dependent population growth 
has obtained much more general criteria
under which a habitat configuration will support a stable polymorphic equilibrium \citep{cantrell1989diffusive},
and determined the habitat configuration that maximizes the probability of establishment \citep{lou2006minimization}.

@@

Consider first a single, isolated poor habitat patch in which no $B$ allele has yet become established. 
We first compute the time scale on which a new $B$ mutations will appear and fix in such a patch,
if the rate of mutation from $b$ to $B$ is $\mu$ per individual per generation.
As we are interested in patches where local adaptation can occur,
we will assume that our patch is larger than the cutoff for local establishment 
mentioned above.

Let $p(x)$ be the probability that a new mutant $B$ allele arising at location $x$
relative to the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approximation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to ignore migration, 
so that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This approximates the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is symmetric about the edge of the patch.
We examine this more generally via exact calculation of $p(x)$ in Appendix \ref{apx:establishment_sims}.

@@

%%%%%%%
\subsection{Simulation methods}
\label{ss:simulations}

First we briefly describe the simulations used below for illustration and validation.
We simulated forward-time dynamics of the number of alleles of each type in a regular grid of demes with fixed size $N$.
Each generation, each individual independently chose to reproduce or not with a probability depending on her type and location in the grid;
locally beneficial alleles were more likely to reproduce.
Each then produced a random, Poisson distributed number of offspring that each independently migrated to demes no more than six steps away
according to a symmetric migration kernel.
Once migrants were distributed, each deme was uniformly resampled back down to $N$ individuals.
(Although we described the simulation in terms of individuals,
we kept track only of total numbers in an equivalent way.)

The reproduction rate in simulations for type $b$ alleles was 30\%; 
this was then multiplied by $1+s$ to get the probability of reproduction for type $B$.
These $s$ are the values reported in the figures, and do not depend on the basic rate of reproduction.
However, to obtain values for $s_p$ and $s_m$ when comparing theory to simulation,
we computed the rate of intrinsic growth, 
i.e.\ the $s$ so that the numbers of $B$ alleles when rare would change by $e^{st}$ after $t$ generations
in the absence of migration.
(The resulting values are close to the first notion of $s$,
but give better agreement with theory,
which uses the second definition.)

To sample lineages, we first simulated forwards in time the population dynamics,
then sampled lineages moving back through time
by, in each generation,
moving each lineage to a new deme with probability proportional to the reverse migration probability
weighted by the number of $B$ alleles in that deme in the previous time step.
If more than one lineage was found in a deme with $n$ $B$ alleles,
then each lineage picked a label uniformly from $1 \ldots n$,
and those picking the same label coalesced.
Since reproduction is Poisson, this correctly samples from the distribution of lineages given the population dynamics.

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, we dissect the process by which an allele transits between the patches,
along the way obtaining other useful information about the genealogy of $B$ alleles.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at location $x$ relative to the patch by $q(x)$,
and assume that the shape of the patch is at least roughly circular.
Following \citet{haldane1948theory} or \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
and show that $q(x)$ decays exponentially for large $|x|$,
with a square-root correction in two dimensions:
\begin{align} \label{eqn:eqfreq}
  q(x) \approx C \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), 
and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients.
These asymptotics fit simulations quite well, as shown in Figure~\ref{fig:sim_occupation_freqs}.
To be clear about the assumptions (especially regarding scaling) implicit here,
we provide a derivation in Appendix~\ref{apx:eqfreq}; 
as well as justification for the asymptotics in Appendix \ref{apx:asymptotics}.
In one dimension, the equation can be solved to give $q(x)$ 
in terms of Jacobi elliptic functions; see Appendix \ref{apx:elliptic_integrals}.
In applications we fit $C$ to the data;
to get concrete numbers from theory we take $C=1$ if necessary.
From simulations, $C$ seems to be not too far from 1 in practice (see below).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \caption{
  \textbf{(A)} 
  Mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  a one-dimensional grid of 201 demes, and
  a two-dimensional grid of $101\times 101$ demes.
  Superimposed in color is the decay predicted by \eqref{eqn:eqfreq},
  with $C$ chosen arbitrarily to give a reasonable fit.
  (The disagreement at long distances is due to boundary effects near the edge of the range.)
  \textbf{(B)} 
  Fluctuations: colored lines are snapshots of allele frequencies at evenly spaced times
  from a one-dimensional grid of 151 demes;
  the solid line shows the cumulative mean occupation frequency across all generations.
  For both plots, the simulation was run for 1000 generations;
  the focal allele is beneficial in the central 10 demes, with $s_p=0.1$,
  and deleterious elsewhere, with $s_m=-0.02$; 
  demes had 1000 individuals in each;
  dispersal was to nearby demes with a mean dispersal distance of $\sigma \approx 1.66$  times the deme spacing.
   }   \label{fig:sim_occupation_freqs}
\end{figure}

This expected frequency $q(x)$ is the \emph{time-averaged} occupation frequency,
i.e.\ the total number of $B$ alleles found across $T$ generations near location $x$, per unit area, divided by $T$.
If, for instance, $q(x)=.01$ and the population density is $\rho=10$ individuals per unit area, 
then in a survey tract of unit area around $x$ we expect to find one individual every 10 generations
-- or, perhaps more likely, 10 individuals every 100 generations.
This points to an important fact: 
close to the original patch, the ``equilibrium frequency'' $q(x)$ describes well the density of the $B$ allele at most times,
but far away from the patch, 
the equilibrium is composed of rare excursions of families of $B$
alleles, interspersed by long periods of absence. 
An example of one of these rare excursions is shown in Figure~\ref{fig:sim_snapshots}.
The relevant time scale on which $B$ alleles migrate between patches is given by the rate of appearance of such families.
To compute the rate of establishment by such migration,
we argue that these families (suitably defined) can be approximated by
independent branching processes whose spatial motion is described by Brownian motion.
Properties of these, combined with equation \eqref{eqn:eqfreq},
will lead us eventually to the expression \eqref{eqn:migrate} for the rate of effective migration,
an approximation that is valid if the distance $R$ is large (at least five times $\sigma/\sqrt{2s_m}$, say).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-snapshots}
  \end{center}
  \caption{
  A segment of the temporal dynamics of the same simulation as Figure~\ref{fig:sim_occupation_freqs}B.
  Space is again shown across the horizontal axis, with time on the vertical axis; 
  demes are colored darker red the greater their allele frequency at the corresponding time.
  The 50\% and 5\% frequency contour lines are in grey,
  and the genealogy of 50 individuals in a rare long-distance excursion is traced back with black lines.
  }   \label{fig:sim_snapshots}
\end{figure}

To make the argument, we decompose the genealogy of $B$ alleles into families in the following way.
First, consider the genealogy of all $B$ alleles that were alive at any time outside of the original patch.  
This is a collection of trees, each rooted at an allele living outside the patch whose parent was born inside the patch.
Next, fix some intermediate distance $r_0$ from the established patch,
and erase from the genealogy every allele that has no ancestors living further away than $r_0$ to the patch.
This fragments the original set of trees into a collection of smaller trees that relate to each other all $B$ alleles living outside of $r_0$ at any time,
and some living inside of $r_0$;
we refer to these trees as ``families''.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated by a collection of independent spatial branching processes
whose members are ignored if they enter the original patch.
(This can be made formal in the limit of large population density, also taking $r_0$ large enough that the chance of reaching the original patch is small.)
This decomposition is illustrated in figure~\ref{fig:branching_decomp}.
In terms of these families, the equilibrium frequency above is then
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        q(x) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation density of a family at $x$ ) } ,
\end{split}
\end{align}
and the quantity we wish to compute,
the effective rate of at which \emph{families of} migrant $B$ alleles establish in a new patch at displacement $x$ from the original,
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(x) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in patch at $x$ ) }
    \end{split}
\end{align}
The quantities on the right-hand side depend implicitly on $r_0$.
All terms except the ``outflux of families'' are calculations with spatial branching processes;
and we can solve for this outflux using the established form for $q(r)$ in \eqref{eqn:gestalt_q}.
Note that $\migrate$ reflects the mean number of establishing migrant families per unit time, rather than individuals
-- a naive calculation using $q(x)$ that assumes migrants are independent would overestimate the effective migration rate.

%%%% %%%%
\subsubsection{The genealogy of migrant families}
